Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-19 ind_files/06_nontechnical_discussion.tex
Minor phrasing changes after proofreading

The equations of section \ref{sec_fun_theorems} generically predict a directional evolutionary force acting on variation in per-capita turnover rate $\tau$ that I call `noise-induced selection' (following previous studies such as~\cite{constable_demographic_2016} and~\cite{week_white_2021}). Noise-induced selection is only seen in finite populations, is seen whenever there is differential turnover rate $\tau$ in the system, and arises due to different types of individuals in the population experiencing a different number of stochastic events (birth and death) in a given time interval. Several specific finite population models, especially in social evolution, illustrate that evolution in finite population can proceed in a direction different from that predicted by the infinite population limit~\citep{parsons_consequences_2010,melbinger_evolutionary_2010, houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,debarre_evolutionary_2016, behar_fluctuations-induced_2016, constable_demographic_2016,abu_awad_effects_2018,parsons_pathogen_2018,mcavoy_public_2018,mcleod_social_2019}. Indeed, in some stochastic models, the outcome of finite population models can be exactly opposite to that of the infinite population limit, a phenomenon sometimes referred to as `reversing the direction of deterministic selection'~\citep{constable_demographic_2016,mcleod_social_2019}. Chapter \ref{chap_unification} explains how the  results of all these studies can be explained using noise-induced selection. Though similar effects have been recognized in several specific models of life-history evolution~\citep{gillespie_natural_1974,kuosmanen_turnover_2022}, social evolution~\citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,constable_demographic_2016,mcleod_social_2019}, and epidemiology~\citep{parsons_pathogen_2018,day_price_2020}, this thesis is, to the best of my knowledge, the first time it has been explicitly systematically derived and formulated in a general, unifying, `model-independent' evolutionary framework.

@@

Being mindful of noise-induced selection is also important for applied fields like conservation and population management which regularly deal with small populations. For example, when trying to increase the population of a hypothetical desired species in a multispecies community, increasing the birth rate is \emph{not} equivalent to reducing the death rate even though both result in an increase in the Malthusian fitness (growth rate) $w_i$. Decreasing the death rate leads to a decrease in $\tau_i$, which leads to positive noise-induced selection, whereas increasing the birth rate leads to an increase in $\tau_i$, which leads to noise-induced selection acting to reduce the abundance of the focal species from the community. If the total community size is small, increasing the birth rate of a species can thus lead to noise-induced selection completely eliminating the focal species from the community despite the fact that we \emph{increased} the growth rate of this species. Indeed, numerical investigations of `burst-death' dynamics in fluctuating viral populations show that increasing Malthusian fitness by boosting the survival rate (\emph{i.e.} reducing the death rate) leads to a greater increase in fixation probability than if an exactly equivalent increase in Malthusian fitness is achieved via increasing the burst rate (\emph{i.e.} increasing the birth rate) of viral particles~\citep{alexander_fixation_2008}.  A recent finite population birth-death model for cancer treatment provides another concrete example of the consequences of the asymmetry between changing birth rates and death rates: Due to the presence of noise-induced selection, the potential of a tumorous growth to adapt to treatments and experience evolutionary rescue depends inversely on the per-capita turnover $\tau_i$ of the constituent cancer cells, with obvious implications for optimal treatment strategies~\citep{raatz_promoting_2023}.

Lastly, noise-induced selection is particular to fluctuating populations and does not occur in models with fixed population sizes\footnote{If $\sum_j x_j$ is a constant, the map $x_i \to x_i/\sum_j x_j$ becomes a linear map and we no longer need It\^o's formula to move from densities to frequencies in the derivation I conduct in Appendix \ref{App_density_to_freq}; Thus, simply dividing equation \eqref{nD_Ito_SDE} by the (now constant) total population size provides the complete dynamics of the system in frequency space: Note that the directional terms in equation \eqref{nD_Ito_SDE} depend only on $\mathbf{A}^-$, which in turn depends only on the fitness $w_i$ and the mutation terms, and this system thus has no noise-induced selection.} such as the neutral Wright-Fisher or Moran models, suggesting that working with such constant population frameworks is not sufficient to accurately capture the dynamics of real populations. Several theoretical studies have pointed out that the empirically common practice of approximating the size of fluctuating populations through a constant `effective population size' (often obtained as the harmonic mean of population size over time) is of limited applicability~\citep{gillespie_natural_1974,sjodin_meaning_2005,parsons_consequences_2010,iizuka_effective_2010,abu_awad_effects_2018,kuosmanen_turnover_2022}. Recent experimental evolution studies have also directly shown that the harmonic mean of population size need not be a good proxy for capturing and/or predicting evolutionary dynamics in fluctuating populations~\citep{chavhan_larger_2019}. In our case, noise-induced selection cannot be captured by any constant effective population size, harmonic mean or otherwise, since noise-induced selection effects also depend on the turnover $\tau_i$ of each type and are thus a property of each \emph{type} in a fluctuating population (and not merely a function of the population size as a whole), meaning that type would demand its own `effective' population size for the entire population. My results thus lend further support to one of the general messages of these previous studies - Approximating fluctuating populations via a simpler constant (effective) population size may inadvertently remove important evolutionary properties of the systems under study and is thus not always a good idea.

At first thought, the idea of an evolutionary force that selects individuals with lower birth and death rates over individuals with higher birth and death rates may be reminiscent of notions in life-history evolution like $r$ vs $K$ selection or selection on the pace of life~\citep{stearns_evolution_1977}. However, it is not possible to conclude whether this similarity reflects some deep principle or whether it is just superficial based solely on the work conducted in this thesis. Models in life-history theory are often primarily concerned with spatiotemporally fluctuating external environments, and thus the stochasticity at the core of life-history models is extrinsic to the population. Ecological frameworks such as modern coexistence theory, which also deal with questions about similar population dynamics and would benefit from a first principles stochastic birth-death formulation, also typically work with fluctuating external environments~\citep{chesson_stabilizing_1982,chesson_multispecies_1994}. I have entirely neglected such extrinsic factors in my formalism. In principle, it is possible to make the birth and death rates \eqref{nD_functional_forms_for_replicator} in my framework also depend on a temporally varying external environment $E(t)$ (whose variation may possibly depend on the population $\mathbf{n}(t)$). Incorporating such a term would ensure that the `ecological feedback' terms in equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} are non-zero, but may also lead to much more complex dynamics. If the variation of the environment $E(t)$ has some associated stochasticity, the complete dynamics of the system would be the result of interactions between two qualitatively different forms of noise --- \emph{extrinsic} noise from the environment, and \emph{intrinsic} noise from the finiteness of the population --- and thus can be rather complicated~\citep{gokhale_eco-evolutionary_2016} and likely difficult to handle analytically. Indeed, experimental studies indicate that (fluctuating) finite population size and fluctuating environments can interact in complex and sometimes counter-intuitive ways, with implications for evolvability and adaptation in spatiotemporally fluctuating environments~\citep{chavhan_larger_2020, chavhan_interplay_2021}. Thus, while integrating the birth-death framework I outline here with ecological ideas such as the pace-of-life syndrome~\citep{mathot_models_2018, wright_life-history_2019} or modern coexistence theory~\citep{chesson_multispecies_1994, johnson_resolving_2022} is biologically appealing, it is likely far from trivial and may present a promising avenue for future work.

~\cite{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `most fundamental' of the lot, and equations like the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in my framework - Equation \eqref{nD_stochastic_RM} is the most fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula, in principle for any moment of the distribution of the quantity $f$ in the population (though this quickly becomes too tedious to actually carry out in practice). If we additionally assume that the quantity $f$ follows a Gaussian distribution in the population, then the mean and variance completely characterize the distribution of $f$, and equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} together specify the complete stochastic dynamics of the system. Note that even in this case, actually solving these equations analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will quickly become impossibly difficult if the birth and death rate functions are complicated. Indeed, previous studies indicate that in high dimensions, evolutionary birth-death models can exhibit a dizzying array of complicated phenomena, including limit cycles and evolutionary chaos~\citep{doebeli_diversity_2017}. However, while this may present impediments to formulating exact solutions to the equations I formulate, the point of these equations is not necessarily to solve them to begin with.

@@

There are many biologically important factors that I have neglected in my formalism as a first order approximation. My framework works with unstructured populations, neglecting any potential effects of groups, age, class, sex, developmental stage, or space\footnote{Of course,  since position in continuous space is just a special case of a quantitative trait, we have technically incorporated space in a very limited sense: The formalism of chapter \ref{chap_infD_processes} can equally well describe a finite population of exactly identical individuals moving through one-dimensional space, and so can technically describe phenomena like range expansion of clonal populations in one dimension. However, I am only noting this as a technicality - Such populations are likely somewhat `evolutionarily boring', since all individuals in the model must always be exactly identical in all aspects other than spatial location.}, all of which can lead to very complex and often surprising dynamics. Explicitly incorporating features such as population structure, sex, and space from first principles in an analytically tractable framework is a formidable task that may need innovative new mathematical and biological arguments, and presents a fantastic opportunity for future studies. I have also neglected any potential complications introduced by genotype-phenotype maps and genetic processes such as dominance, epistasis, and pleiotropy since they can (in principle) be incorporated into the definition of the per-capita birth and death rates in \eqref{nD_functional_forms_for_replicator}.

>>>>

The equations of section \ref{sec_fun_theorems} generically predict a directional evolutionary force acting on variation in per-capita turnover rate $\tau$ that I call `noise-induced selection' (following previous studies such as~\cite{constable_demographic_2016} and~\cite{week_white_2021}). Noise-induced selection is only seen in finite populations, is seen whenever there is differential turnover rate $\tau$ in the system, and arises due to different types of individuals in the population experiencing a different number of stochastic events (birth and death) in a given time interval. Several specific finite population models, especially in social evolution, illustrate that evolution in finite population can proceed in a direction different from that predicted by the infinite population limit~\citep{parsons_consequences_2010,melbinger_evolutionary_2010, houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,debarre_evolutionary_2016, behar_fluctuations-induced_2016, constable_demographic_2016,veller_drift-induced_2017,abu_awad_effects_2018,parsons_pathogen_2018,mcavoy_public_2018,mcleod_social_2019}. Indeed, in some stochastic models, the outcome of finite population models can be exactly opposite to that of the infinite population limit, a phenomenon sometimes referred to as `reversing the direction of deterministic selection'~\citep{constable_demographic_2016,mcleod_social_2019}. Chapter \ref{chap_unification} explains how the  results of all these studies can be explained using noise-induced selection. Though similar effects have been recognized in several specific models of evolutionary ecology~\citep{gillespie_natural_1974,veller_drift-induced_2017,kuosmanen_turnover_2022}, social evolution~\citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,constable_demographic_2016,mcleod_social_2019}, and epidemiology~\citep{parsons_pathogen_2018,day_price_2020}, this thesis is, to the best of my knowledge, the first time it has been explicitly systematically derived and formulated in a general, unifying, `model-independent' evolutionary framework.

@@

Being mindful of noise-induced selection is also important for applied fields that regularly deal with manipulating small populations, such as conservation and population management. For example, when trying to increase the population of a hypothetical desired species in a multi-species community, increasing the birth rate is \emph{not} equivalent to reducing the death rate even though both result in an increase in the Malthusian fitness (growth rate) $w_i$. Decreasing the death rate leads to a decrease in $\tau_i$, which is favored by noise-induced selection, whereas increasing the birth rate leads to an increase in $\tau_i$, which leads to noise-induced selection acting to reduce the abundance of the focal species from the community. If the total community size is small, increasing the birth rate of a species can thus lead to noise-induced selection completely eliminating the focal species from the community despite the fact that we \emph{increased} the growth rate of this species. Indeed, numerical investigations of `burst-death' dynamics in fluctuating viral populations show that increasing Malthusian fitness by boosting the survival rate (\emph{i.e.} reducing the death rate) leads to a greater increase in fixation probability than if an exactly equivalent increase in Malthusian fitness is achieved via increasing the burst rate (\emph{i.e.} increasing the birth rate) of viral particles~\citep{alexander_fixation_2008}.  A recent finite population birth-death model for cancer treatment provides another concrete example of the consequences of the asymmetry between changing birth rates and death rates: Due to the presence of noise-induced selection, the potential of a tumorous growth to adapt to treatments and experience evolutionary rescue depends inversely on the per-capita turnover $\tau_i$ of the constituent cancer cells, with obvious implications for optimal treatment strategies~\citep{raatz_promoting_2023}.

Lastly, noise-induced selection is particular to fluctuating populations and does not occur in models with fixed population sizes\footnote{If $\sum_j x_j$ is a constant, the map $x_i \to x_i/\sum_j x_j$ becomes a linear map and we no longer need It\^o's formula to move from densities to frequencies in the derivation I conduct in Appendix \ref{App_density_to_freq}; Thus, simply dividing equation \eqref{nD_Ito_SDE} by the (now constant) total population size provides the complete dynamics of the system in frequency space: Note that the directional terms in equation \eqref{nD_Ito_SDE} depend only on $\mathbf{A}^-$, which in turn depends only on the fitness $w_i$ and the mutation terms, and this system thus has no noise-induced selection.} such as the neutral Wright-Fisher or Moran models, suggesting that working with such constant population frameworks is not sufficient to accurately capture the dynamics of real populations. Several theoretical studies have pointed out that the empirically common practice of approximating the size of fluctuating populations through a constant `effective population size' (often obtained as the harmonic mean of population size over time) is of limited applicability~\citep{gillespie_natural_1974,sjodin_meaning_2005,parsons_consequences_2010,iizuka_effective_2010,abu_awad_effects_2018,kuosmanen_turnover_2022}. Recent experimental evolution studies have also directly shown that the harmonic mean of population size need not be a good proxy for capturing and/or predicting evolutionary dynamics in fluctuating populations~\citep{chavhan_larger_2019}. In our case, noise-induced selection cannot be captured by any constant effective population size, harmonic mean or otherwise, since noise-induced selection effects also depend on the turnover $\tau_i$ of each type and are thus a property of each \emph{type} in a fluctuating population (and not merely a function of the population size as a whole), meaning that each type would demand its own effective population size for the entire population. My results thus lend further support to one of the general messages of these previous studies - Approximating fluctuating populations via a simpler constant (effective) population size requires some careful justification, since it may inadvertently remove important evolutionary properties of the systems under study.

At first thought, the idea of an evolutionary force that selects individuals with lower birth and death rates over individuals with higher birth and death rates may be reminiscent of notions in life-history evolution like $r$ vs $K$ selection or selection on the pace of life~\citep{stearns_evolution_1977}. However, it is not possible to conclude whether this similarity reflects some deep principle or whether it is just superficial based solely on the work conducted in this thesis. Models in life-history theory are often primarily concerned with spatiotemporally fluctuating external environments, and thus the stochasticity at the core of life-history models is extrinsic to the population. Ecological frameworks such as modern coexistence theory, which also deal with questions about similar population dynamics and would benefit from a first principles stochastic birth-death formulation, also typically work with fluctuating external environments~\citep{chesson_stabilizing_1982,chesson_multispecies_1994,barabas_chessons_2018, johnson_resolving_2022}. I have entirely neglected such extrinsic factors in my formalism. In principle, it is possible to make the birth and death rates \eqref{nD_functional_forms_for_replicator} in my framework also depend on a temporally varying external environment $\mathbf{E}(t)$ (whose variation may possibly depend on the population $\mathbf{n}(t)$). Incorporating such a term would ensure that the `ecological feedback' terms in equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} are non-zero, but may also lead to much more complex dynamics. If the variation of the environment $\mathbf{E}(t)$ has some associated stochasticity, the complete dynamics of the system would be the result of interactions between two qualitatively different forms of noise --- \emph{extrinsic} noise from the environment, and \emph{intrinsic} noise from the finiteness of the population --- and thus can be rather complicated~\citep{gokhale_eco-evolutionary_2016} and likely difficult to handle analytically. Indeed, experimental studies indicate that (fluctuating) finite population size and fluctuating environments can interact in complex and sometimes counter-intuitive ways, with implications for evolvability and adaptation in spatiotemporally fluctuating environments~\citep{chavhan_larger_2020, chavhan_interplay_2021}. Thus, while integrating the birth-death framework I outline here with ecological ideas such as the pace-of-life syndrome~\citep{mathot_models_2018, wright_life-history_2019} or modern coexistence theory~\citep{barabas_chessons_2018, johnson_resolving_2022} is biologically appealing, it is likely far from trivial and may present a promising avenue for future work.

~\cite{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `most fundamental' equation, and equations like the dynamically sufficient version of the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in my framework - Equation \eqref{nD_stochastic_RM} is the most fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula, in principle for any moment of the distribution of the quantity $f$ in the population (though this quickly becomes too tedious to actually carry out in practice). If we additionally assume that the quantity $f$ follows a Gaussian distribution in the population, then the mean and variance completely characterize the distribution of $f$, and equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} together specify the complete stochastic dynamics of the system. Note that even in this case, actually solving these equations analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will quickly become impossibly difficult if the birth and death rate functions are complicated. Indeed, previous studies indicate that in high dimensions, evolutionary birth-death models can exhibit a dizzying array of complicated behavior, including limit cycles and chaotic trajectories~\citep{doebeli_diversity_2017}. However, while this may present impediments to formulating exact solutions to the equations I formulate, the point of these equations is not necessarily to solve them to begin with.

@@

My formalism here is only a first step, and there are many biologically important factors that I have neglected in this thesis. My framework works with unstructured populations, neglecting any potential effects of groups, age, class, sex, developmental stage, or space\footnote{Of course,  since position in continuous space is just a special case of a quantitative trait, we have technically incorporated space in a very limited sense: The formalism of chapter \ref{chap_infD_processes} can equally well describe a finite population of exactly identical individuals moving through one-dimensional space, and so can technically describe phenomena like range expansion of clonal populations in one dimension. However, I am only noting this as a technicality - Such populations are likely somewhat `evolutionarily boring', since all individuals in the model must always be exactly identical in all aspects other than spatial location.}, all of which can lead to very complex and often surprising dynamics. Explicitly incorporating features such as population structure, sex, and space from first principles in an analytically tractable framework is a formidable task that may need innovative new mathematical and biological arguments, and presents a fantastic opportunity for future studies. I have also neglected any potential complications introduced by genotype-phenotype maps and genetic processes such as dominance, epistasis, and pleiotropy since my assumptions on the functional forms of the birth and death rates (\eqref{nD_functional_forms_for_replicator} or \eqref{BD_for_kimura}) should be general enough to incoporate these effects (in principle) while analyzing specific models.

### 2023-03-19 ind_files/99_Appendix_A.tex
Minor phrasing changes after proofreading

Consider a one-dimensional real It\^{o} process given by $dX_t = \mu(X_t,t)dt + \sigma(X_t,t)dW_t$ that takes values in a subset $\Omega \subseteq \mathbb{R}$ and admits a probability density function $P(x,t)$ on $\Omega \times [0,\infty)$. Let $f:\mathbb{R}\to\mathbb{R}$ be an arbitrary $C^2(\mathbb{R})$ function. By It\^{o}'s formula, we have:

@@

We will now use integration by parts to further evaluate the two parts $M(x,t)$ and $N(x,t)$. Recall that the general formula for integration by parts is given by:
    \int\limits_{\Omega}u_{x_i}vdx = -\int\limits_{\Omega}uv_{x_i}dx + \int\limits_{\partial\Omega}uv\gamma_{i}dS(x)
where subscript indicates differentiation and $\gamma$ is the unit outward normal. In our case, assuming that $P(x,t) \equiv 0$ on $\partial \Omega$, the boundary term (second term of the RHS) vanishes and we can use integration by parts once on $M(x,t)$ to obtain

>>>>

Consider a one-dimensional real It\^{o} process given by $dX_t = \mu(X_t,t)dt + \sigma(X_t,t)dW_t$ that takes values in an open set $\Omega \subseteq \mathbb{R}$ and admits a probability density function $P(x,t)$ with compact support\footnote{For our biological context you can think of this as just meaning that very extreme values in $\Omega$ are so unlikely that we disallow them entirely} in $\Omega$ at all times $t \in [0,\infty)$. Let $f:\mathbb{R}\to\mathbb{R}$ be an arbitrary $C^2(\mathbb{R})$ function. By It\^{o}'s formula, we have:

@@

We will now use integration by parts to further evaluate the two parts $M(x,t)$ and $N(x,t)$. Since $P(\cdot, t)$ has compact support within the open set $\Omega \subset \mathbb{R}$ for any given $t > 0$, the function $P(x,t) \equiv 0$ on $\partial \Omega$ and the boundary term of the formula vanishes. Thus, we can use integration by parts once on $M(x,t)$ to obtain
